Experiences in Modelling a Microcontroller Instruction Set Using B

### 2012-01-19 papers/ArticleZ80/source/paper.tex

The B method~\cite{Abrial} supports the construction of safety systems
models by generating proof obligation that must be verified to guarantee its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined up to the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. \cite{Dantas_SBMF08} proposed
an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

This work presents a formal modell of the instruction set of the Z80 
microcontroller~\cite{Z80_manual}. This microcontroller
was chosen because of several factors: it contains essential and common
concepts in microcontrollers and until microprocessors; it has extensive
documentation available; and it was widely used and is still in use. 
A first Z80 modell presented in \cite{Valerio_SBMF09} was verified completely,
but it uses constructs that are not well supported to animate. 
This modelling was changed to support animation in ProB\cite{proB}.
Basically, the new modelling\footnote{The interested reader in more
details is invited to visit our repository at: http://code.google.com/p/b2asm/.}
changed elements represented by infinite sets and adjusted the implications
of this change. Using the responsibility division mechanisms provided by B, auxiliary
libraries of basic modules were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers;besides the new Z80 model, a theoretic case study in 
petroleum production test system was developed using Z80 assembly language
to analyse the code simulation and verification process.

The formal model of instruction set provides many benefits. The animation of 
Z80 model allows to simulate the execution assembly programs including
support for interrupts and input and output instructions. Other possible uses
of a formal model of a microcontroller instruction set include documentation,
the construction of simulators, and can be possibly the starting point of a 
verification effort for the actual implementation of a Z80 design. Moreover, 
the model of the instruction set could be instrumented with non-functional
aspects, such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine.
Two aspects of these formalization are of particular importance to us:
provide a more solid documentation artifact for microcontrollers, 
build a reference model for a formal compilation approach in B.

\subsection{Problems in formalizing instruction set}

In general, the manuals of microcontrollers can be more suited for
developers in assembly language. Many times, each instruction of microcontrollers
is shown in its official manual using textual description, math description, examples,
encoding, number of cycles and other information.Some times, the instruction descriptions
have an own notation, some informations are organized in different sections of document 
and the textual description are not standardized. This does not allow the interpretation by parser.
% b) The case of Z80
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description. So, this description can add
some ambiguities or even mistakes. The official manual has also several problems that were 
identified over time~\cite{UndocumentedZ80}. The Z80 is a microcontroller that  has been
tested and used for many years. This intensive use and test facilitated to find several
errors in the description of the instructions in its manual. Several of these problems
have been reported in a technical report \cite{UndocumentedZ80}. Some problems are inaccurate
information, for example the description of the action of an instruction on the flag
register, partial information and distributed on different pages of manual.

% in the Z80 manual happens was that,

%A simple example of error is in some official documentation saying that a group of instructions leaves the \textit{CF} flag unaffected and the \textit{NF}
%flag is not always set, but may also be reset \cite{UndocumentedZ80}.
%% \begin{itemize} \item Some input and output instructionsThe
%% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some
%% official documentation says they leave it unaffected, important!) and the NF
%% flag is not always set but may also be reset (see 4.3 for exact operation)
%% \cite{UndocumentedZ80}. Este erro he descrito na phegina (274 do documento e
%% 294 do pdf) do manual oficial do Z80 (Z80_CPU_adotado.pdf) Citar apenas esse
%% exemplo abaixo! \item
%%when a determined interruption (NMI) is accepted, the bit IFF1 is not   
%%copied to bit IFF2~\cite{UndocumentedZ80}. 
%%
%This example can cause serious problems for programmers.
%% Este erro he descrito na pagina (22 do documento e 26do pdf) do manual
%% official do Z80 (Z80_CPU_adotado.pdf) Only IFF1 is reset.
%This error and others 7 different errors were previously identified by~\cite{UndocumentedZ80}. So to help the 
%construction of Z80 B model, other references~\cite{UndocumentedZ80,Simulator_z80}
%were also consulted.

@@

%Finally, the main goal of this project, though, is to provide a basis for the
%generation of software artefacts at the assembly level, those are amenable to 
%refinement verification within the B method.

%\subparagraph{Report structure}

This work is focused on the presentation of modelling of assembly instruction
set, including elementary libraries to describe hardware aspects; and the
modelling and simulation of a case study of an embedded verified software that
can be used in tanks of the petroleum industry.

This paper is structured as it follows. %Section~\ref{sec:B_method} provides a short introduction to the B method. 
Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. %Section ~\ref{sec:Verification} provides some information on the proof effort needed to analyse the presented models.
Section ~\ref{sec:studycase} explains the building and verifying assembly-level refinements. Related works are discussed in
section~\ref{sec:relatedworks}. Finally, the last section is devoted to the conclusions.
%
%\section{Introduction to the B Method}
%\label{sec:B_method}
%
%The B method for software development~\cite{Abrial} is based on the B Abstract
%Machine Notation (AMN) for the formal specification of functional requirements and the use of formally proved refinements up to a
%specification sufficiently concrete that programming code can be automatically
%generated from it. Its mathematical basis consists of first order logic, integer
%arithmetic and set theory, and its corresponding constructs are similar to those
%of the Z notation~\cite{Spivey96}.
%
%A B specification is structured in modules. A module defines a set of valid
%states, including a set of initial states, and operations that may provoke a
%transition among states. The design process starts with a module with a
%so-called functional model of the system under development. In this initial
%modelling stage, the B method requires that the user proves that, in a machine,
%all its initial states are valid, and operations do not define
%transitions from valid states to invalid states.
%
%
%Essentially, a B module contains two main parts: a header and the available
%operations. Figure~\ref{fig:maqB} has a very basic example. The clause
%\MACHINE  has the name of module.  The next two clauses, respectively,
%reference external modules and create an instance of an external module.  The
%\VARIABLES clause declares the name of the variables that compose the
%state of the machine. Next, the \INVARIANT clause defines the valid states,
%specifying  the type and other restrictions on the variables. The \INITIALISATION specifies
%the initial states. Finally, operations correspond to the transitions among
%states of the machine.
%
%
%\begin{figure}[h]
%  % \begin{small}
%\begin{center}
%\begin{sloppypar}
%
%  $\begin{array}{lcl}
%    \begin{array}[t]{l}
%      \MACHINE \quad  \textit{micro} \\
%      \SEES \quad \textit{TYPES}, \textit{ALU} \\
%      \INCLUDES  \quad \textit{MEMORY} \\
%      \VARIABLES \quad    \textit{pc}  \\
%      \INVARIANT\\ 
%      \quad \textit{pc} \in \textit{INSTRUCTION} 
%    \end{array}
%    & \hspace*{0cm} &
%    \begin{array}[t]{l}
%      \INITIALISATION  \textit{pc} := 0 \\
%      \OPERATIONS\\
%      \textit{JMP} \textit{( jump )} = \\
%      \quad \PRE \textit{jump} \in \textit{INSTRUCTION}\\ 
%      \quad \THEN \textit{pc} := \textit{jump}\\  
%      \quad \END \\
%      \END
%    \end{array}
%  \end{array}
%  $
% \end{sloppypar} 
%\end{center}
%\caption{A very basic B machine.}
%\label{fig:maqB}
%\end{figure}

@@

First, bits are modelled as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits and it is
defined as:

@@

\subsection{Modelling bytes and bit vectors of length 16}
\label{subsec:HardwareLibrary3}

@@

The $\textit{BYTE\_INDEX}$ is the domain of the functions modelling bytes. It starts at 1 to obey a
definition of sequences from B. However, it is common in hardware architectures to start indexing from
zero. The definition $\textit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\textit{BYTE}$ type is as a specialized type of $\textit{BIT\_VECTOR}$, but it has a size
limit. Other specific definitions are provided to facilitate further modelling: the type $\textit{BV16}$
is created for bit vector of length 16 in a similar way.

@@

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modelling library.  For example, the function $\textit{bv\_to\_nat}$ from bit vector arithmetic is
specialized to $\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can defined as follows:

@@

@@

\bf MACHINE

\hspace*{0.15in}\it Z80

\bf INCLUDES

\hspace*{0.10in}\it MEMORY

\bf SEES

\hspace*{0.10in}\it ALU, \it BIT\_DEFINITION, \it BIT\_VECTOR\_DEFINITION,

\hspace*{0.10in}\it BYTE\_DEFINITION, \it BV16\_DEFINITION,

\hspace*{0.10in}\it UCHAR\_DEFINITION, \it SCHAR\_DEFINITION,

\hspace*{0.10in}\it SSHORT\_DEFINITION ,\it USHORT\_DEFINITION
\end{sloppypar}

\subsection{Modelling registers}

The Z80 CPU includes alternative set of accumulator, flag and general registers. The CPU contains a stack
pointer ($\textit{sp}$), program counter ($\textit{pc}$), two index registers ($\textit{ix}$ and $\textit{iy}$), an
interrupt register ($\textit{i\_}$), a refresh register ($\textit{r\_}$), two bits ($\textit{iff1}$,
$\textit{iff2}$) used to control the interruptions, a pair of bits to define the interruption mode ($\textit{im}$)
and the input and output ports ($\textit{i\_o\_ports}$). Below, its definitions are represented by
\textit{INVARIANT}.

\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$

\hspace*{0.10in}\it pc  $\in$  \it INSTRUCTION  $\land$  \it sp  $\in$  \it BV16  $\land$  \it ix  $\in$  \it BV16  $\land$  \it iy  $\in$  \it BV16  $\land$

\hspace*{0.10in}\it i\_  $\in$  \it BYTE  $\land$  \it r\_ $\in$  \it BYTE  $\land$

\hspace*{0.10in}\it iff1  $\in$  \it BIT  $\land$ \hspace*{0.10in}\it iff2  $\in$  \it BIT  $\land$

\hspace*{0.10in}\it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$

\hspace*{0.10in}\it i\_o\_ports  $\in$  \it BYTE  $\fun$  \it BYTE
\end{sloppypar}

% Features of microcontroller
  The internal registers contain 176 bits of read/write memory that are represented by
identifiers used as parameters in the instructions . It includes two sets of six general purpose
registers which may be used individually as 8-bit registers or as 16-bit register pairs.  The working
registers are represented by variable $\textit{rgs8}$. The domain of $\textit{rgs8}$
($\textit{id\_regs8}$) is a set formed by identifiers of registers of 8 bits. These registers are called
main register set (\textit{a0,b0,c0,d0,e0,f0,h0,l0}) and alternate register set
(\textit{a\_0,b\_0,c\_0,d\_0,e\_0,f\_0,h\_0,l\_0}). These registers can be accessed in pairs, forming
16-bits, resulting in another set of identifiers of 16-bits registers, named $\textit{id\_reg16}$.

	\bf SETS\\
	\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0 \rm , \it f0 \rm , \it f\_0 \rm , \it a\_0 \rm ,\\
	\hspace*{0.93in}\it b0 \rm , \it c0 \rm , \it b\_0 \rm , \it c\_0\rm,\\
	\hspace*{0.93in}\it d0 \rm , \it e0 \rm , \it d\_0 \rm , \it e\_0\rm,\\
	\hspace*{0.93in}\it h0 \rm , \it l0 \rm , \it h\_0 \rm , \it l\_0\};\\
	\hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{ \it BC \rm , \it DE \rm , \it HL \rm , \it SP \rm , \it AF \rm \}

@@

\subsection{Modelling the actions and instructions}

Each instruction is represented by a B operation in the module Z80. The main
module (\textit{Z80}) has three categories of operations: a category represents the 
instructions of microcontrollers, a second category represents the input and output of data 
(shown in ~\ref{sec:modellingIO}) and the last category represents the external
actions (shown in~\ref{sec:externalactions}). A simple example of instruction category is
$\textit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The tools B does not allow
to use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module and it receives
a address memory and its new memory value. It also increments the program
counter ($\textit{pc}$) and update the refresh register ($\textit{r\_}$).
The other instructions have a similar structure.

@@

\subsection{Modelling the input/output instructions}
\label{sec:modellingIO}
The Z80 has an extensive set of input and output (I/O) instructions and 256 ports for
devices. This model can transfer data blocks between the I/O devices and any
of the internal registers or memory address.

@@

\subsection{Modelling the external actions}
\label{sec:externalactions}

The external actions change the state of the microcontroller, for example, refreshing the I/O ports
and interruptions request. The external actions are also modelled by operations and they are named with the
prefix ``$ext\_$'' and followed by the name of action. There are four external
actions: $ext\_update\_io\_ports$, $ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The
$ext\_update\_io\_ports$ just change the state of I/O port, see.

@@

The interruptions and the \textit{reset} action can change the state of
program counter. Theses actions are modelled by B operations and its main
effects are shown below\footnote{Some definitions of constants: $\textit{sp\_minus\_two}$ holds the value of stack pointer minus 2,
%=$\textit{dec\_BV16(dec\_BV16(sp))}$
 $\textit{sp\_minus\_one}$ is the value of stack pointer minus 1,
%=$\textit{dec\_BV16(sp)}$
$\textit{pc\_high}$ holds the most significant 8 bits and
$\textit{pc\_low}$ holds the least significant 8 bits.}.

@@

The modelling of instruction set, interruptions and input/output ports become the animation interesting.
The animation in ProB version is not so fast and practical as a simulation, for example in \cite{Simulator_z80},
because ProB works manipulating logic and math concepts that can have infinite sets and non determinism, then ProB requires more processing, specially when it works analysing  complex expressions or very large expressions.
However, a user of ProB with this modelling can: execute manually a sequence of Z80's instructions that represent a program and view the visited states of ports, memory and registers, it is important to debug the effect action on the state space computed; create a tracing of the states; analyse constants, functions and expressions; and search possible violations.

@@

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object of the pilot project developed to analyse a petroleum production test
system in each oil field. Its modelling was developed in according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modelling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

@@

There are in the literature several
approaches~\cite{BHDL_2003,DBLP:journals/entcs/EvansG08,DBLP:conf/fmco/Leuschel08,DBLP:conf/asm/Wright08}
to model hardware and the virtual machines using B. In these works, the B
has been used successfully to model the operational semantic. However the cost of modelling is still expensive.

These recent works \cite{Marc20113,springerlink:Yuan2011} have a similar approach using Event B. 
However, our work use the B method, that seems more appropriated to software development, because it has
implementable language defined, called B0, and tools to convert the models to a programming language.

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus some aspects were not
modelled in our work such as the execution time of the instructions.
Also we did not consider the micro architecture of the hardware as the scope of
our work does not include hardware verification.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modelling of the instruction set of
microcontroller using the B method. During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. The designer fixed the found problems in
documentation and also developed a case study: a verified embedded software. This
case study was interesting to analyse the technique of formal verification up to
assembly level.

@@

>>>>

The B method~\cite{Abrial} supports the formal development of software
from a specification of functional requirements down to an imperative
implementation of those requirements, amenable to synthesis to a
programming language such as C or Ada. However, the synthesis step is
not amenable to proof of correctness, and~\cite{Dantas_SBMF08} proposed
an approach to extend the scope of the B method up to the assembly
level language. One key component of this approach is to build, within
the framework of the B method, formal models of the instruction set of
such assembly languages.

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
documentation available; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it uses constructs that make it impossible to animate.

The new model supports animation in ProB\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of microcontroller model. Such
library has many definitions of concepts common in microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using Z80 assembly language to analyse the code simulation
and verification process.

Providing the formal model of an instruction set yields many
benefits. The animation of Z80 model allows to simulate the execution
assembly programs including support for interrupts and input and
output instructions. Other possible uses of a formal model of a
microcontroller instruction set include documentation, the
construction of simulators, and can be possibly the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of these formalization
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers, build a reference model for a formal
compilation approach in B.

\subsection{Problems in formalizing instruction set}
% Compact version

In general, the manuals of microcontrollers can be more suited for \marginpar{DD: o título da sub-seção não condiz com seu conteúdo. Não cita nenhum problema na formalização. Não vejo muito bem o que ele traz de novo que já não foi dito anteriormente.}
developers in assembly language. Many times, each instruction of
microcontrollers is shown in its official manual using textual
description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. This
does not allow the interpretation by parser.  The manuals also have
semi-formal and informal descriptions, these descriptions may make
mistakes.  The official manual of the microcontroller Z80 has also
several problems that were identified over
time~\cite{UndocumentedZ80}.

@@

%In general, the manuals of microcontrollers can be more suited for
%developers in assembly language. Many times, each instruction of microcontrollers
%is shown in its official manual using textual description, math description, examples,
%encoding, number of cycles and other information.Some times, the instruction descriptions
%have an own notation, some informations are organized in different sections of document 
%and the textual description are not standardized. This does not allow the interpretation by parser.
%% b) The case of Z80
%For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users
%want to know the action of an instruction and its effects on the flag register then he needs
%to read an informal textual description. So, this description can add
%some ambiguities or even mistakes. The official manual has also several problems that were 
%identified over time~\cite{UndocumentedZ80}. The Z80 is a microcontroller that  has been
%tested and used for many years. This intensive use and test facilitated to find several
%errors in the description of the instructions in its manual. Several of these problems
%have been reported in a technical report \cite{UndocumentedZ80}. Some problems are inaccurate
%information, for example the description of the action of an instruction on the flag
%register, partial information and distributed on different pages of manual.
%
%
%% in the Z80 manual happens was that,
%
%%A simple example of error is in some official documentation saying that a group of instructions leaves the \textit{CF} flag unaffected and the \textit{NF}
%%flag is not always set, but may also be reset \cite{UndocumentedZ80}.
%%% \begin{itemize} \item Some input and output instructionsThe
%%% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some
%%% official documentation says they leave it unaffected, important!) and the NF
%%% flag is not always set but may also be reset (see 4.3 for exact operation)
%%% \cite{UndocumentedZ80}. Este erro he descrito na phegina (274 do documento e
%%% 294 do pdf) do manual oficial do Z80 (Z80_CPU_adotado.pdf) Citar apenas esse
%%% exemplo abaixo! \item
%%%when a determined interruption (NMI) is accepted, the bit IFF1 is not   
%%%copied to bit IFF2~\cite{UndocumentedZ80}. 
%%%
%%This example can cause serious problems for programmers.
%%% Este erro he descrito na pagina (22 do documento e 26do pdf) do manual
%%% official do Z80 (Z80_CPU_adotado.pdf) Only IFF1 is reset.
%%This error and others 7 different errors were previously identified by~\cite{UndocumentedZ80}. So to help the 
%%construction of Z80 B model, other references~\cite{UndocumentedZ80,Simulator_z80}
%%were also consulted.
%
%A solution to avoid ambiguity and inconsistencies is to specify formally the
%assembly instruction set, this creates a pattern of representation and validates
%properties of instructions. The formal model also restricts the definitions to
%correct typing and uses only well-defined expressions. Furthermore, the developer
%can use with B method different levels of abstraction and add more specific details
%in refined models.
% 
%Finally, the main goal of this project, though, is to provide a basis for the
%generation of software artefacts at the assembly level, those are amenable to 
%refinement verification within the B method.

%\subparagraph{Report structure}
%
%This work is focused on the presentation of modelling of assembly instruction
%set, including elementary libraries to describe hardware aspects; and the
%modelling and simulation of a case study of an embedded verified software that
%can be used in tanks of the petroleum industry.

This paper is structured as it follows.  Section~\ref{sec:models}
presents the elementary libraries and the modeling of some elements
common to microcontrollers.  Section~\ref{sec:z80} presents the B
model of the Z80 instruction set.  Related works are discussed in
section~\ref{sec:relatedworks}. Finally, the last section is devoted
to the conclusions.

%
%\section{Introduction to the B Method}
%\label{sec:B_method}
%
%The B method for software development~\cite{Abrial} is based on the B Abstract
%Machine Notation (AMN) for the formal specification of functional requirements and the use of formally proved refinements up to a
%specification sufficiently concrete that programming code can be automatically
%generated from it. Its mathematical basis consists of first order logic, integer
%arithmetic and set theory, and its corresponding constructs are similar to those
%of the Z notation~\cite{Spivey96}.
%
%A B specification is structured in modules. A module defines a set of valid
%states, including a set of initial states, and operations that may provoke a
%transition among states. The design process starts with a module with a
%so-called functional model of the system under development. In this initial
%modelling stage, the B method requires that the user proves that, in a machine,
%all its initial states are valid, and operations do not define
%transitions from valid states to invalid states.
%
%
%Essentially, a B module contains two main parts: a header and the available
%operations. Figure~\ref{fig:maqB} has a very basic example. The clause
%\MACHINE  has the name of module.  The next two clauses, respectively,
%reference external modules and create an instance of an external module.  The
%\VARIABLES clause declares the name of the variables that compose the
%state of the machine. Next, the \INVARIANT clause defines the valid states,
%specifying  the type and other restrictions on the variables. The \INITIALISATION specifies
%the initial states. Finally, operations correspond to the transitions among
%states of the machine.
%
%
%\begin{figure}[h]
%  % \begin{small}
%\begin{center}
%\begin{sloppypar}
%
%  $\begin{array}{lcl}
%    \begin{array}[t]{l}
%      \MACHINE \quad  \textit{micro} \\
%      \SEES \quad \textit{TYPES}, \textit{ALU} \\
%      \INCLUDES  \quad \textit{MEMORY} \\
%      \VARIABLES \quad    \textit{pc}  \\
%      \INVARIANT\\ 
%      \quad \textit{pc} \in \textit{INSTRUCTION} 
%    \end{array}
%    & \hspace*{0cm} &
%    \begin{array}[t]{l}
%      \INITIALISATION  \textit{pc} := 0 \\
%      \OPERATIONS\\
%      \textit{JMP} \textit{( jump )} = \\
%      \quad \PRE \textit{jump} \in \textit{INSTRUCTION}\\ 
%      \quad \THEN \textit{pc} := \textit{jump}\\  
%      \quad \END \\
%      \END
%    \end{array}
%  \end{array}
%  $
% \end{sloppypar} 
%\end{center}
%\caption{A very basic B machine.}
%\label{fig:maqB}
%\end{figure}

@@

First, bits are modeled as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits and it is
defined as:

@@

\subsection{Modeling bytes and bit vectors of length 16}
\label{subsec:HardwareLibrary3}

@@

The $\textit{BYTE\_INDEX}$ is the domain of the functions modeling bytes. It starts at 1 to obey a
definition of sequences from B. However, it is common in hardware architectures to start indexing from
zero. The definition $\textit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\textit{BYTE}$ type is as a specialized type of $\textit{BIT\_VECTOR}$, but it has a size
limit. Other specific definitions are provided to facilitate further modeling: the type $\textit{BV16}$
is created for bit vector of length 16 in a similar way.

@@

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modeling library.  For example, the function $\textit{bv\_to\_nat}$ from bit vector arithmetic is
specialized to $\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can defined as follows:

@@

	\begin{center}
	\begin{sloppypar}

	  \begin{tabbing}
	    \begin{array}[t]{l}
		\bf MACHINE\\
		\hspace*{0.15in}\it Z80\\
		\bf INCLUDES\\
		\hspace*{0.10in}\it MEMORY
	    \end{array}
	     \hspace*{1cm} 
	    \begin{array}[t]{l}
	    \bf SEES\\
		\hspace*{0.10in}\it ALU, \it BIT\_DEFINITION, \it BIT\_VECTOR\_DEFINITION,\\
		\hspace*{0.10in}\it BYTE\_DEFINITION, \it BV16\_DEFINITION,\\
		\hspace*{0.10in}\it UCHAR\_DEFINITION, \it SCHAR\_DEFINITION,\\
		\hspace*{0.10in}\it SSHORT\_DEFINITION ,\it USHORT\_DEFINITION\\
	    \end{array}
	  \end{tabbing}

	 \end{sloppypar} 
	\end{center}

\subsection{Modeling registers}

The Z80 CPU includes alternative set of accumulator, flag and general registers. The CPU contains a stack
pointer ($\textit{sp}$), program counter ($\textit{pc}$), two index registers ($\textit{ix}$ and $\textit{iy}$), an
interrupt register ($\textit{i\_}$), a refresh register ($\textit{r\_}$), two bits ($\textit{iff1}$,
$\textit{iff2}$) used to control the interruptions, a pair of bits to define the interruption mode ($\textit{im}$)
and the input and output ports ($\textit{i\_o\_ports}$). These definitions are represented by
\textit{INVARIANT}.

\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ \it pc  $\in$  \it INSTRUCTION  $\land$  \it sp  $\in$  \it BV16 $\land$  \it ix  $\in$  \it BV16  $\land$

\hspace*{0.10in}\it iy  $\in$  \it BV16  $\land$ \it i\_  $\in$  \it BYTE  $\land$  \it r\_ $\in$  \it BYTE  $\land$ \it iff1  $\in$  \it BIT  $\land$ \it iff2  $\in$  \it BIT  $\land$

\hspace*{0.10in}\it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$  \it i\_o\_ports  $\in$  \it BYTE  $\fun$  \it BYTE
\end{sloppypar}

% Features of microcontroller
  The internal registers contain 176 bits of read/write memory that are represented by
identifiers used as parameters in the instructions . It includes two sets of six general purpose
registers which may be used individually as 8-bit registers or as 16-bit register pairs.  The working
registers are represented by variable $\textit{rgs8}$. The domain of $\textit{rgs8}$
($\textit{id\_regs8}$) is a set formed by identifiers of registers of 8 bits. These registers are called
main register set (\textit{a0,b0,c0,d0,e0,f0,h0,l0}) and alternate register set
(\textit{a\_0,b\_0,c\_0,d\_0,e\_0,f\_0,h\_0,l\_0}). These registers can be accessed in pairs, forming
16-bits, resulting in another set of identifiers of 16-bits registers, named $\textit{id\_reg16}$. These sets are represented below.

@@

	\bf SETS\\
	\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0 \rm , \it f0 \rm ,\it b0 \rm , \it c0 \rm ,\it d0 \rm , \it e0 \rm ,\it h0 \rm , \it l0 \rm ,\it a\_0 \rm , \it f\_0 \rm ,\it b\_0 \rm , \it c\_0 \rm ,\it d\_0 \rm , \it e\_0 \rm ,\it h\_0 \rm , \it l\_0 \rm \};\\
	\hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{ \it BC \rm , \it DE \rm , \it HL \rm , \it SP \rm , \it AF \rm \}

@@

\subsection{Modeling the actions and instructions}

Each instruction is represented by a B operation in the module Z80. The main
module (\textit{Z80}) has three categories of operations: a category represents the 
instructions of microcontrollers, a second category represents the input and output of data 
(shown in ~\ref{sec:modelingIO}) and the last category represents the external
actions (shown in~\ref{sec:externalactions}). A simple example of instruction category is
$\textit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The tools B does not allow
to use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module and it receives
a address memory and its new memory value. It also increments the program
counter ($\textit{pc}$) and update the refresh register ($\textit{r\_}$).
The other instructions have a similar structure.

@@

\subsection{Modeling the input/output instructions}
\label{sec:modelingIO}
The Z80 has an extensive set of input and output (I/O) instructions and 256 ports for
devices. This model can transfer data blocks between the I/O devices and any
of the internal registers or memory address.

@@

\subsection{Modeling the external actions}
\label{sec:externalactions}

The external actions change the state of the microcontroller, for example, refreshing the I/O ports
and interruptions request. The external actions are also modeled by operations and they are named with the
prefix ``$ext\_$'' and followed by the name of action. There are four external
actions: $ext\_update\_io\_ports$, $ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The
$ext\_update\_io\_ports$ just change the state of I/O port, see.

@@

The interruptions and the \textit{reset} action can change the state of
program counter. Theses actions are modeled by B operations and its main
effects are shown below\footnote{Some definitions of constants: $\textit{sp\_minus\_two}$ holds the value of stack pointer minus 2,
%=$\textit{dec\_BV16(dec\_BV16(sp))}$
 $\textit{sp\_minus\_one}$ is the value of stack pointer minus 1,
%=$\textit{dec\_BV16(sp)}$
$\textit{pc\_high}$ holds the most significant 8 bits and
$\textit{pc\_low}$ holds the least significant 8 bits.}.

@@

The modeling of instruction set, interruptions and input/output ports become the animation interesting.
The animation in ProB version is not so fast and practical as a simulation, for example in \cite{Simulator_z80},
because ProB works manipulating logic and math concepts that can have infinite sets and non determinism, then ProB requires more processing, specially when it works analysing  complex expressions or very large expressions.
However, a user of ProB with this modeling can: execute manually a sequence of Z80's instructions that represent a program and view the visited states of ports, memory and registers, it is important to debug the effect action on the state space computed; create a tracing of the states; analyse constants, functions and expressions; and search possible violations.

@@

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object of the pilot project developed to analyse a petroleum production test
system in each oil field. Its modeling was developed in according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modeling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

@@

There are in the literature several similar approaches to model hardware and instruction set using B.
The paper \cite{springerlink:Yuan2011} reports a method to specify, design and construct sound and complete instruction set model by stepwise refinement and formal proof using the formal method Event-B. It also discuss about desirable properties of instruction set model. However, our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that seems more appropriated to software development, because it has implementable language defined, 
called B0, and tools to convert the models to a programming language.
In \cite{Marc20113} is present an related experience on using B in the design of secure micro-controllers. It tries to show the feasibility of such a technique for high confidence trustful devices. The work \cite{Subotic2010} describes a similar approach with MIPS CPU architecture and it shows how the CPU is formally specified and implemented using the B method.

% 
%There are in the literature several
%approaches~\cite{BHDL_2003,DBLP:journals/entcs/EvansG08,DBLP:conf/fmco/Leuschel08,DBLP:conf/asm/Wright08}
%to model hardware and the virtual machines using B. In these works, the B
%has been used successfully to model the operational semantic. However the cost of modeling is still expensive. 
%
%These recent works \cite{Marc20113,springerlink:Yuan2011} have a similar approach using Event B. 
%However, our work use the B method, that seems more appropriated to software development, because it has
%implementable language defined, called B0, and tools to convert the models to a programming language.
%
%
%The main motivation of our research is the development of verified
%software up to the assembly level, which requires specifying the
%semantics of the underlying hardware. Thus some aspects were not
%modeled in our work such as the execution time of the instructions.
%Also we did not consider the micro architecture of the hardware as the scope of
%our work does not include hardware verification.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the instruction set of
microcontroller using the B method. During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. The designer fixed the found problems in
documentation and also developed a case study: a verified embedded software. This
case study was interesting to analyse the technique of formal verification up to
assembly level.

@@

\subsubsection{Acknowledgements:} This experiment could not have succeeded without the interesting work of Michael Leuschel who developed several improvements in the ProB and provided important aid to enable the animation of the Z80 modeling.

% This work was partially supported by INES (www.ines.org.br),
%funded by CNPq grant 573964/2008-4 and by CNPq grants 553597/2008-6, 550946/2007-
%1, and 620132/2008-6
